---
title: "Prevalence of depression, anxiety and PTSD among Ukrainian Refugees in host countries: systematic review and meta-analysis"
author: "Tamara"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
---
```{r setup, include=FALSE}
library(readxl)
library(dplyr)
library(metafor)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
# Introduction

Context of the displacement: Russian full-scale invasion in February 2022
Mental health risks in forced migration: find stats on mental health risks in forced migration, particularly in context of war refugees - how this typically differs from say economical migrants; find prevalence of mental health conditions in general population too
Exisiting evidence: individual studies and previous meta-data from other war refugees groups (if available?)
Knowledge gap: No previous meta-analysis focused on Ukrainian refugees
Additional topics to introduce: anxiety, depression and PTSD and validated assessment tools

# Methods

## Search strategy

Databases searched: PubMed, EuropePMC

Date of searches: 06/01/2026

Search terms:

a) PubMed:
(Ukrainian OR Ukraine) AND
(refugee* OR displaced) AND
(anxiety OR depression OR PTSD OR "post-traumatic stress") AND
(prevalence OR symptom*)

b) EuropePMC:
(Ukrainian OR Ukraine) AND refugee AND
(anxiety OR depression OR PTSD) AND prevalence

Filter: Open access

Records identified (total): 505
a) PubMed: 45
b) EuropePMC: 460

After duplicates removal: 485
Title/ abstract screening: 485
Full text screening: 41
Included: 11

## Eligibility criteria

Inclusion criteria:

- Population: Ukrainian refugees displaced after Feb 2022
- Outcomes: depression, anxiety, PTSD
- Study designs: cross‑sectional, cohort baseline
- Validated self‑report mental‑health questionnaires: PHQ‑9, PHQ‑8, PHQ‑4, GAD‑7, BAI, ITQ, IES‑R, PCL‑5, CRIES‑13

Exclusion criteria:

- Studies inside Ukraine
- Qualitative studies
- Non‑validated self‑report mental‑health questionnaires
- No extractable prevalence data; no means/SD

## Data extraction

Study info:

study_id	first_author	year	title	journal	country_host	study_design	sampling_method	n_total	age_mean	age_sd	percent_female	population_notes	time_since_displacement_months	refugee_status_definition	risk_of_bias_sampling	risk_of_bias_measurement	risk_of_bias_reporting	overall_risk_of_bias	notes

Outcomes info:

study_id	outcome_domain	instrument	cutoff_value	cutoff_definition	n_assessed	n_above_cutoff	prevalence_percent	mean_score	sd_score	response_rate	notes_on_measurement

## Statistical Analysis

Effect size: logit-transformed proportions (PLO)  
Model: random-effects (Paule–Mandel)  
Heterogeneity: I², τ², Q  
Sensitivity analyses: leave-one-out, RoB exclusion  
No meta-regression due to <10 studies per outcome

# Results

```{r load-data}
study   <- read_excel("C:/Users/tamci/OneDrive/Desktop/R projects/Meta-analysis/meta-analysis extraction.xlsx", sheet = "Study")
outcome <- read_excel("C:/Users/tamci/OneDrive/Desktop/R projects/Meta-analysis/meta-analysis extraction.xlsx", sheet = "Outcome")

names(study)
names(outcome)
```

```{r prevalence}
#Looking at prevalence
dat_prev <- outcome %>%
  filter(!is.na(n_assessed),
         !is.na(n_above_cutoff)) %>%
  left_join(study, by = "study_id") %>%
  mutate(
    prop = n_above_cutoff / n_assessed
  )

#Cross-checking with excel sheet
dat_prev %>%
  select(study_id, outcome_domain, instrument,
         n_assessed, n_above_cutoff, prop, prevalence_percent)
```
## Depression meta-analysis
```{r depression}
#Random-effect meta-analysis for depression
#Depression prevalence
dat_dep <- dat_prev %>%
  filter(outcome_domain == "depression")

es_dep <- escalc(measure = "PLO",
                 xi = n_above_cutoff,
                 ni = n_assessed,
                 data = dat_dep)

res_dep <- rma(yi, vi, data = es_dep, method = "PM")  # or method = "REML"

summary(res_dep)

#Back-transform pooled estimate and prediction interval to prevalence
predict(res_dep, transf = transf.ilogit, digits = 3)
```

Forest plot depression
```{r forest-plot-depression}
# Forest plot for depression
forest(res_dep,
       slab = dat_dep$study_id,
       transf = transf.ilogit,
       refline = predict(res_dep, transf = transf.ilogit)$pred,
       xlab = "Prevalence (proportion)",
       main = "Depression prevalence among Ukrainian refugees")
```

## Anxiety meta-analysis
```{r anxiety}
#Random-effect meta-analysis for anxiety
dat_anx <- dat_prev %>%
  filter(outcome_domain == "anxiety")

es_anx <- escalc(measure = "PLO",
                 xi = n_above_cutoff,
                 ni = n_assessed,
                 data = dat_anx)

res_anx <- rma(yi, vi, data = es_anx, method = "PM")
summary(res_anx)
predict(res_anx, transf = transf.ilogit, digits = 3)
```
Forest plot anxiety
```{r forest-plot-anxiety}
# Forest plot for anxiety
forest(res_anx,
       slab = dat_anx$study_id,
       transf = transf.ilogit,
       refline = predict(res_anx, transf = transf.ilogit)$pred,
       xlab = "Prevalence (proportion)",
       main = "Anxiety prevalence among Ukrainian refugees")
```

## PTSD meta-analysis
```{r PTSD}
#Random-effect meta-analysis for PTSD
dat_ptsd <- dat_prev %>%
  filter(outcome_domain == "PTSD")

es_ptsd <- escalc(measure = "PLO",
                  xi = n_above_cutoff,
                  ni = n_assessed,
                  data = dat_ptsd)

res_ptsd <- rma(yi, vi, data = es_ptsd, method = "PM")
summary(res_ptsd)
predict(res_ptsd, transf = transf.ilogit, digits = 3)
```
Forest plot PTSD
```{r forest-plot-PTSD}
# Forest plot for PTSD
forest(res_ptsd,
       slab = dat_ptsd$study_id,
       transf = transf.ilogit,
       refline = predict(res_ptsd, transf = transf.ilogit)$pred,
       xlab = "Prevalence (proportion)",
       main = "PTSD prevalence among Ukrainian refugees")
```
Sensitivity Analyses
```{r sensitivity-analyses}
# Sensitivity analyses
leave1out(res_dep)
leave1out(res_anx)
leave1out(res_ptsd)
```
Subgroup Analysis by Host Country
```{r subgroup-analysis-by-host-country}
study$host_country <- c(
  "Germany",
  "Poland",
  "Multiple",
  "Germany",
  "Germany",
  "Czechia",
  "Denmark",
  "Lithuania",
  "Poland",
  "Hungary",
  "Czechia",
  "Switzerland"
)

dat_prev <- outcome %>%
  filter(!is.na(n_assessed),
         !is.na(n_above_cutoff)) %>%
  left_join(study, by = "study_id") %>%
  mutate(prop = n_above_cutoff / n_assessed)

dat_dep$host_country  <- study$host_country[match(dat_dep$study_id, study$study_id)]
dat_anx$host_country  <- study$host_country[match(dat_anx$study_id, study$study_id)]
dat_ptsd$host_country <- study$host_country[match(dat_ptsd$study_id, study$study_id)]

run_subgroup_models <- function(data) {
  
  # Splitting dataset into a list of dataframes, one per host country
  subgroups <- split(data, data$host_country)
  
  # For each subgroup, computing effect sizes and fitting a random-effects model
  models <- lapply(subgroups, function(df) {
    
    # Calculating logit-transformed proportions (PLO)
    es <- escalc(
      measure = "PLO",
      xi = n_above_cutoff,
      ni = n_assessed,
      data = df
    )
    
    # Fitting random-effects model using Paule-Mandel estimator
    rma(yi, vi, data = es, method = "PM")
  })
  
  return(models)
}

# Running subgroup models for each outcome
dep_models  <- run_subgroup_models(dat_dep)
anx_models  <- run_subgroup_models(dat_anx)
ptsd_models <- run_subgroup_models(dat_ptsd)

# Extracting pooled prevalence (back-transformed)
extract_prevalence <- function(models) {
  lapply(models, function(m) {
    predict(m, transf = transf.ilogit)   # converting logit back to proportion
  })
}

# Extracting prevalence estimates by country
dep_prev_by_country  <- extract_prevalence(dep_models)
anx_prev_by_country  <- extract_prevalence(anx_models)
ptsd_prev_by_country <- extract_prevalence(ptsd_models)

# Printing results
dep_prev_by_country
anx_prev_by_country
ptsd_prev_by_country

# Forest plots 
plot_subgroup_forests <- function(data, outcome_name) {
  
  # Splitting dataset by host country
  subgroups <- split(data, data$host_country)
  
  # Looping through each country and generate a forest plot
  for (country in names(subgroups)) {
    
    df <- subgroups[[country]]
    
    # Computing effect sizes
    es <- escalc(
      measure = "PLO",
      xi = n_above_cutoff,
      ni = n_assessed,
      data = df
    )
    
    # Fitting random-effects model
    model <- rma(yi, vi, data = es, method = "PM")
    
    print(
      forest(
        model,
        slab = df$study_id,              # study labels
        transf = transf.ilogit,          # back-transform to proportions
        main = paste(outcome_name, "prevalence –", country)
      )
    )
  }
}

# Generating subgroup forest plots (these now print automatically)
plot_subgroup_forests(dat_dep,  "Depression")
plot_subgroup_forests(dat_anx,  "Anxiety")
plot_subgroup_forests(dat_ptsd, "PTSD")
```